---
title: "Investigating the Impacts of Ocean Acidfication on Zooplankton and Krill Biovolumes"
subtitle: "Part II"
format:
  html:
    toc: true
fig-pos: H
---

```{r setup, echo = F}
knitr::opts_chunk$set(echo = F)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
```

```{r}
# Libraries
library(readr)
library(here)
library(gt)
library(dplyr)
library(ggplot2)
library(tidyr)
library(gam)
library(mgcv)
```

# Introduction


# Problems of Interest

Our goal is to study the impact of ocean acidification on zooplankton and krill biovolumes off of the California coast. First, we want to conduct a cross-comparison of bottle data and biological data to assess the amount of spatial and temporal overlap present. Using co-located measurements, we wish to model the effects of ocean acidification, using carbonate chemistry and oceanographic variables, on zooplankton and krill abundance. In addition, we are interested in exploring how pH and related environmental factors affect the abundance of calcifying versus non-calcifying species. 

# Data

## Bottle Data^[Bottle Database: <https://calcofi.org/data/oceanographic-data/bottle-database/>]

CalCOFI has collected environmental and hydrographic data for over 70 years during their quarterly cruises to their CalCOFI stations. We used the merged data from Part I which contains oceanographic and carbonate chemistry data as well as computed CO2SYS values for variables such as $pH$, $CO_3$, and $\Omega_{aragonite}$. This merged dataset we are working with is called merged_bottle_co2sys.csv. One important thing to note is that there is a large temporal gap between the years 2002-2007.

There are three datasets that we are focusing on for our biological data: (1) CalCOFI NOAA Zooplankton Volume, (2) BTEDB (Krill) Abundances, and (3) PRPOOS Data (for Zooplankton Calcifiers/Non-Calcifiers Abundance). The zooplankton and krill biovolume data are obtained using net tows (Bongo and/or Pairovet) at each standard CalCOFI station. The PRPOOS data is also obtained using a net tow but rather than sampling at station, it is conducted during transits between stations. These three datasets have each been merged with the bottle data to create zoop_data/zooplankton_pH.csv, krill_data/CV_merged_krill.csv, and PRPOOS/prpoos_summary.csv, respectively. 

## CalCOFI NOAA Zooplankton Volume^[CalCOFI Zooplankton Volume Database: <https://oceanview.pfeg.noaa.gov/erddap/tabledap/erdCalCOFIzoovol.html>]:
The zooplankton biovolume data measures the amount of “plankton” (the small and microscopic organisms floating in the sea, consisting chiefly of diatoms, protozoans, small crustaceans, and the eggs and larval stages of larger animals) in the volume of sea water sampled. In particular, we are interested in the variables `total_plankton` and `small_plankton`.

## BTEDB (Krill) Abundances^[BTEDB (Krill Volume) Data: <https://portal.edirepository.org/nis/mapbrowse?packageid=knb-lter-cce.313.1>]:

The krill dataset provides information on krill abundance from the Brinton and Townsend Euphausiid Database (BTEDB). The samples collected include species such as *Euphausia pacifica*, *Nematoscelis difficilis*, and *Thysanoessa spinifera*, with individuals categorized by size and developmental phase (e.g., calyptopis, furcilia, juvenile, adult).

## PRPOOS (Calcifiers/Non-Calcifiers)^[PRPOOS (Zooplankton Calcifiers/Non-Calcifiers Volume) Zooscan Database: <https://oceaninformatics.ucsd.edu/zooscandb/>]:
The PRPOOS (Planktonic Rate Processes in Oligotrophic Ocean Systems) dataset contains abundance and estimated biomass values for various zooplankton taxa, which can be categorized into calcifying and non-calcifying groups. The calcifying taxa are defined as *byrozoan larvae*, *pteropoda heteropoda*, *ostracods*, and *rhizaria*; the remaining taxa are considered non-calcifying. 


# Methods 


# Results

```{r}

df <- data.frame(
  Rows = c("# of Observations", "# of Stations", "Start Year", "End Year"),
  Bottle = c(4125, 51, 1983, 2021),
  zoop_og = c(45310, 3172, 1951, 2023),
  zoop_merge = c(434, 28, 1987, 2021),
  krill_og = c(7482, 411, 1951, 2019),
  krill_merge = c("FILL", "FILL", "FILL", "FILL"),
  prpoos_og = c(1384, 20, 2005, 2025),
  prpoos_merge = c("FILL", "FILL", "FILL", "FILL"),
  stringsAsFactors = FALSE
)

df %>%
  gt(rowname_col = "Rows") %>%
  tab_header(title = "Merged Data") %>%
  cols_label(
    Bottle = "Bottle*",
    zoop_og = "Original", zoop_merge = "Merged",
    krill_og = "Original", krill_merge = "Merged",
    prpoos_og = "Original", prpoos_merge = "Merged"
  ) %>%
  tab_spanner(label = "NOAA Zooplankton", columns = c(zoop_og, zoop_merge)) %>%
  tab_spanner(label = "BTEDB Krill", columns = c(krill_og, krill_merge)) %>%
  tab_spanner(label = "PRPOOS Calcifiers", columns = c(prpoos_og, prpoos_merge)) %>%
  tab_source_note(source_note = "*Note temporal gap from 2002-2007.")

```


## Zooplankton Models

### Total Plankton

```{r}
zooplankton <- read_csv("../data/zoop_data/zooplankton_pH.csv")

total_zooplankton <- zooplankton %>% 
  dplyr::select(CTDTEMP_ITS90, Salinity_PSS78, DIC, TA, total_plankton, pHout, Latitude, Longitude, Year_UTC, Month_UTC, Station_ID, pCO2in, RFin, CO3in, OmegaCAin, OmegaARin) %>%
  drop_na() %>%
  mutate(log_total_plankton = log1p(total_plankton)) %>%
  dplyr::select(-total_plankton) %>%
  mutate(Station_ID = as.factor(Station_ID))

gam_model4 <- gam(
  log_total_plankton ~ 
    te(Longitude, Latitude) +
                  s(pHout) +
                  s(TA) + s(DIC) +
                  s(Salinity_PSS78) +
                  s(CTDTEMP_ITS90) +
                  s(Year_UTC) + s(Month_UTC) + s(Station_ID, bs = "re"),
  data = total_zooplankton,
)
summary(gam_model4)
AIC(gam_model4)
plot(gam_model4)

plot(gam_model4, select = 1, shade = TRUE, seWithMean = TRUE, scheme = 2)

par(mfrow = c(4, 2), mar = c(4, 4, 2, 1)) 
for (i in 2:9) {
  plot(gam_model4, select = i, shade = TRUE)
}
```

### Small Plankton

```{r}
small_zooplankton <- zooplankton %>% 
  dplyr::select(CTDTEMP_ITS90, Salinity_PSS78, DIC, TA, small_plankton, pHout, Latitude, Longitude, Year_UTC, Month_UTC, Station_ID, pCO2in, RFin, CO3in, OmegaCAin, OmegaARin) %>%
  drop_na() %>%
  mutate(log_small_plankton = log1p(small_plankton)) %>%
  dplyr::select(-small_plankton) %>%
  mutate(Station_ID = as.factor(Station_ID))

small_gam_mod2 <- gam(
  log_small_plankton ~ 
    te(Longitude, Latitude) +
                  s(pHout) +
                  s(TA) + s(DIC) +
                  s(Salinity_PSS78) +
                  s(CTDTEMP_ITS90) +
                  s(Year_UTC) + s(Month_UTC) + s(Station_ID, bs = "re"),
  data = small_zooplankton,
)

summary(small_gam_mod2)
AIC(small_gam_mod2)
plot(small_gam_mod2)
plot(small_gam_mod2, select = 1, shade = TRUE, seWithMean = TRUE, scheme = 2)
#plot(small_gam_mod2, page = 1, shade = TRUE, seWithMean = TRUE)

par(mfrow = c(4, 2), mar = c(4, 4, 2, 1)) 
for (i in 2:9) {
  plot(small_gam_mod2, select = i, shade = TRUE)
}


par(mfrow = c(1, 1))

```

## Krill Models

### PRPOOS data

```{r}
library(lubridate)
library(mgcv)
library(ggplot2)
library(dplyr)
library(maps)
library(stringr)
# read in data
prpoos_summary <- read_csv("../data/PRPOOS/prpoos_summary_namecleaned.csv")
```


```{r}
abundance_vars <- names(prpoos_summary)[str_detect(names(prpoos_summary), "Abundance$")]

# Example for one species, you can wrap in loop later
species <- "bryozoan_larvaeAbundance"

# Fit the GAM
formula_text <- as.formula(paste("log1p(", species, ") ~ s(Longitude, Latitude, k = 15) + s(pH_mean) + s(OmegaCA_mean) + s(CO3_mean) + s(TA_mean) + s(DIC_mean) + s(RF_mean) + s(OmegaAR_mean)"))
gam_model_bl <- gam(formula_text, data = prpoos_summary)

# Make a prediction grid over your spatial extent
lon_seq <- seq(min(prpoos_summary$Longitude) - 3,
               max(prpoos_summary$Longitude) + 3,
               length.out = 200)
lat_seq <- seq(min(prpoos_summary$Latitude) - 3,
               max(prpoos_summary$Latitude) + 3,
               length.out = 200)
grid <- expand.grid(Longitude = lon_seq, Latitude = lat_seq)

# Use mean values for non-spatial covariates
grid$pH_mean <- mean(prpoos_summary$pH_mean, na.rm = TRUE)
grid$OmegaCA_mean <- mean(prpoos_summary$OmegaCA_mean, na.rm = TRUE)
grid$CO3_mean <- mean(prpoos_summary$CO3_mean, na.rm = TRUE)
grid$TA_mean <- mean(prpoos_summary$TA_mean, na.rm = TRUE)
grid$DIC_mean <- mean(prpoos_summary$DIC_mean, na.rm = TRUE)
grid$RF_mean <- mean(prpoos_summary$RF_mean, na.rm = TRUE)
grid$OmegaAR_mean <- mean(prpoos_summary$OmegaAR_mean, na.rm = TRUE)

# Predict from GAM
grid$fit <- predict(gam_model_bl, newdata = grid)

# World map data
world <- map_data("world")

# Plot with ggplot2
ggplot() +
  geom_raster(data = grid, aes(x = Longitude, y = Latitude, fill = fit), alpha = 0.8) +  # transparency
  borders("world", colour = "black") +
  coord_quickmap(xlim = range(prpoos_summary$Longitude) + c(-2, 2),
                 ylim = range(prpoos_summary$Latitude) + c(-2, 2)) +
  scale_fill_viridis_c(name = paste("log1p(", species, ")", sep = "")) +
  labs(title = paste("GAM Spatial Smooth for", gsub("Abundance$", "", species)),
       x = "Longitude", y = "Latitude") +
  geom_point(data = prpoos_summary, aes(x = Longitude, y = Latitude), 
             shape = 21, fill = "white", color = "black", size = 2, stroke = 0.3) + 

  theme_minimal()
summary(gam_model_bl)
AIC(gam_model_bl)
plot(gam_model_bl)

```

# Summary of Findings


# Future Work


# References

