# CO2SYS
## data/merged_bottle_co2sys.csv
- Filter <= 212 m depth for net tows
- Summary Stats for `pH` and `OmegaCA`
```{r}
# Read in dataset
co2sys <- read_csv(here::here("data/merged_bottle_co2sys.csv"))
# Filter depth
co2sys <- co2sys %>%
  dplyr::select(Year_UTC, Month_UTC, Station_ID, Depth, pHin, OmegaCAin, CO3in) %>%
  filter(Depth <= 212)
```

# PRPOOS
## data/PRPOOS
```{r}
# Read in datasets
byrozoan_larvae <- read_csv(here::here("data/PRPOOS/byrozoan_larvae.csv"), skip = 2)
pteropoda_heteropoda <- read_csv(here::here("data/PRPOOS/pteropoda_heteropoda.csv"), skip = 2)
ostracods <- read_csv(here::here("data/PRPOOS/ostracods.csv"), skip = 2)
rhizaria <- read_csv(here::here("data/PRPOOS/rhizaria.csv"), skip = 2)

### all taxa
prpoos <- read_csv(here::here("data/PRPOOS/PRPOOS_all.csv"), skip = 2)
```

```{r}
# Preprocessing datasets
## Combine `Line` and `Station` (in that order) to create `Station_ID` variable
## Extract Month and Year from `Station date` into separate variables

### byrozoan_larvae
byrozoan_larvae$Station_ID <- paste(
  sprintf('%05.1f', byrozoan_larvae$Line),
  sprintf('%05.1f', byrozoan_larvae$Station),
  sep = ' '
)
byrozoan_larvae <- byrozoan_larvae %>%
  relocate(Station_ID, .after = Line)
byrozoan_larvae <- byrozoan_larvae %>%
  mutate(
    Year_UTC = year(`Station date`),
    Month_UTC = month(`Station date`),
    Day_UTC = day(`Station date`),
    .after = `Station date`
  ) %>%
  rename (Abundance = `Abundance (No. per m2)`)

### pteropoda_heteropoda
pteropoda_heteropoda$Station_ID <- paste(
  sprintf('%05.1f', pteropoda_heteropoda$Line),
  sprintf('%05.1f', pteropoda_heteropoda$Station),
  sep = ' '
)
pteropoda_heteropoda <- pteropoda_heteropoda %>%
  relocate(Station_ID, .after = Line)
pteropoda_heteropoda <- pteropoda_heteropoda %>%
  mutate(
    Year_UTC = year(`Station date`),
    Month_UTC = month(`Station date`),
    Day_UTC = day(`Station date`),
    .after = `Station date`
  ) %>%
  rename (Abundance = `Abundance (No. per m2)`)

### ostracods
ostracods$Station_ID <- paste(
  sprintf('%05.1f', ostracods$Line),
  sprintf('%05.1f', ostracods$Station),
  sep = ' '
)
ostracods <- ostracods %>%
  relocate(Station_ID, .after = Line)
ostracods <- ostracods %>%
  mutate(
    Year_UTC = year(`Station date`),
    Month_UTC = month(`Station date`),
    Day_UTC = day(`Station date`),
    .after = `Station date`
  ) %>%
  rename (Abundance = `Abundance (No. per m2)`)

### rhizaria
rhizaria$Station_ID <- paste(
  sprintf('%05.1f', rhizaria$Line),
  sprintf('%05.1f', rhizaria$Station),
  sep = ' '
)
rhizaria <- rhizaria %>%
  relocate(Station_ID, .after = Line)
rhizaria <- rhizaria %>%
  mutate(
    Year_UTC = year(`Station date`),
    Month_UTC = month(`Station date`),
    Day_UTC = day(`Station date`),
    .after = `Station date`
  ) %>%
  rename (Abundance = `Abundance (No. per m2)`)

### all taxa
prpoos$Station_ID <- paste(
  sprintf('%05.1f', prpoos$Station),
  sprintf('%05.1f', prpoos$Line),
  sep = ' '
)
prpoos <- prpoos %>%
  relocate(Station_ID, .after = Line)
prpoos <- prpoos %>%
  mutate(
    Year_UTC = year(`Station date`),
    Month_UTC = month(`Station date`),
    Day_UTC = day(`Station date`),
    .after = `Station date`
  ) %>%
  rename (`appendicularia Abundance (No. per m2)` = `Day or Night`)
```

```{r}
# Merge

### byrozoan_larvae
byrozoan_larvae_merge <- co2sys %>%
  left_join(byrozoan_larvae, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()
write_csv(byrozoan_larvae_merge, here::here("data/PRPOOS/byrozoan_larvae_merge.csv"))

### pteropoda_heteropoda
pteropoda_heteropoda_merge <- co2sys %>%
  left_join(pteropoda_heteropoda, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()
write_csv(pteropoda_heteropoda_merge, here::here("data/PRPOOS/pteropoda_heteropoda_merge.csv"))

### ostracods
ostracods_merge <- co2sys %>%
  left_join(ostracods, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()
write_csv(ostracods_merge, here::here("data/PRPOOS/ostracods_merge.csv"))

### rhizaria
rhizaria_merge <- co2sys %>%
  left_join(rhizaria, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()
write_csv(rhizaria_merge, here::here("data/PRPOOS/rhizaria_merge.csv"))

### all
prpoos_merge <- co2sys %>%
  left_join(prpoos, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()
write_csv(prpoos_merge, here::here("data/PRPOOS/prpoos_merge.csv"))
```

```{r}
# Generate summary stats for pH, total calcium, and carbonate

### byrozoan_larvae
byrozoan_larvae_summary <- byrozoan_larvae_merge %>%
  group_by(Station_ID, Month_UTC, Year_UTC, Abundance) %>%
  summarise(
    pH_min = min(pHin, na.rm = TRUE),
    pH_max = max(pHin, na.rm = TRUE),
    pH_mean = mean(pHin, na.rm = TRUE),
    pH_median = median(pHin, na.rm = TRUE),
    pH_var = var(pHin, na.rm = TRUE),
    pH_range = max(pHin, na.rm = TRUE) - min(pHin, na.rm = TRUE),
    depth_at_pH_min = Depth[which.min(pHin)],
    depth_at_pH_max = Depth[which.max(pHin)],
    
    OmegaCA_min = min(OmegaCAin, na.rm = TRUE),
    OmegaCA_max = max(OmegaCAin, na.rm = TRUE),
    OmegaCA_mean = mean(OmegaCAin, na.rm = TRUE),
    OmegaCA_median = median(OmegaCAin, na.rm = TRUE),
    OmegaCA_var = var(OmegaCAin, na.rm = TRUE),
    OmegaCA_range = max(OmegaCAin, na.rm = TRUE) - min(OmegaCAin, na.rm = TRUE),
    depth_at_OmegaCA_min = Depth[which.min(OmegaCAin)],
    depth_at_OmegaCA_max = Depth[which.max(OmegaCAin)],
    .groups = "drop"
  )
write_csv(byrozoan_larvae_summary, here::here("data/PRPOOS/byrozoan_larvae_summary.csv"))

### pteropoda_heteropoda
pteropoda_heteropoda_summary <- pteropoda_heteropoda_merge %>%
  group_by(Station_ID, Month_UTC, Year_UTC, Abundance) %>%
  summarise(
    pH_min = min(pHin, na.rm = TRUE),
    pH_max = max(pHin, na.rm = TRUE),
    pH_mean = mean(pHin, na.rm = TRUE),
    pH_median = median(pHin, na.rm = TRUE),
    pH_var = var(pHin, na.rm = TRUE),
    pH_range = max(pHin, na.rm = TRUE) - min(pHin, na.rm = TRUE),
    depth_at_pH_min = Depth[which.min(pHin)],
    depth_at_pH_max = Depth[which.max(pHin)],
    
    OmegaCA_min = min(OmegaCAin, na.rm = TRUE),
    OmegaCA_max = max(OmegaCAin, na.rm = TRUE),
    OmegaCA_mean = mean(OmegaCAin, na.rm = TRUE),
    OmegaCA_median = median(OmegaCAin, na.rm = TRUE),
    OmegaCA_var = var(OmegaCAin, na.rm = TRUE),
    OmegaCA_range = max(OmegaCAin, na.rm = TRUE) - min(OmegaCAin, na.rm = TRUE),
    depth_at_OmegaCA_min = Depth[which.min(OmegaCAin)],
    depth_at_OmegaCA_max = Depth[which.max(OmegaCAin)],
    .groups = "drop"
  )
write_csv(pteropoda_heteropoda_summary, here::here("data/PRPOOS/pteropoda_heteropoda_summary.csv"))

### ostracods
ostracods_summary <- ostracods_merge %>%
  group_by(Station_ID, Month_UTC, Year_UTC, Abundance) %>%
  summarise(
    pH_min = min(pHin, na.rm = TRUE),
    pH_max = max(pHin, na.rm = TRUE),
    pH_mean = mean(pHin, na.rm = TRUE),
    pH_median = median(pHin, na.rm = TRUE),
    pH_var = var(pHin, na.rm = TRUE),
    pH_range = max(pHin, na.rm = TRUE) - min(pHin, na.rm = TRUE),
    depth_at_pH_min = Depth[which.min(pHin)],
    depth_at_pH_max = Depth[which.max(pHin)],
    
    OmegaCA_min = min(OmegaCAin, na.rm = TRUE),
    OmegaCA_max = max(OmegaCAin, na.rm = TRUE),
    OmegaCA_mean = mean(OmegaCAin, na.rm = TRUE),
    OmegaCA_median = median(OmegaCAin, na.rm = TRUE),
    OmegaCA_var = var(OmegaCAin, na.rm = TRUE),
    OmegaCA_range = max(OmegaCAin, na.rm = TRUE) - min(OmegaCAin, na.rm = TRUE),
    depth_at_OmegaCA_min = Depth[which.min(OmegaCAin)],
    depth_at_OmegaCA_max = Depth[which.max(OmegaCAin)],
    .groups = "drop"
  )
write_csv(ostracods_summary, here::here("data/PRPOOS/ostracods_summary.csv"))

### rhizaria
rhizaria_summary <- rhizaria_merge %>%
  group_by(Station_ID, Month_UTC, Year_UTC, Abundance) %>%
  summarise(
    pH_min = min(pHin, na.rm = TRUE),
    pH_max = max(pHin, na.rm = TRUE),
    pH_mean = mean(pHin, na.rm = TRUE),
    pH_median = median(pHin, na.rm = TRUE),
    pH_var = var(pHin),
    pH_range = max(pHin, na.rm = TRUE) - min(pHin, na.rm = TRUE),
    depth_at_pH_min = Depth[which.min(pHin)],
    depth_at_pH_max = Depth[which.max(pHin)],
    
    OmegaCA_min = min(OmegaCAin, na.rm = TRUE),
    OmegaCA_max = max(OmegaCAin, na.rm = TRUE),
    OmegaCA_mean = mean(OmegaCAin, na.rm = TRUE),
    OmegaCA_median = median(OmegaCAin, na.rm = TRUE),
    OmegaCA_var = var(OmegaCAin, na.rm = TRUE),
    OmegaCA_range = max(OmegaCAin, na.rm = TRUE) - min(OmegaCAin, na.rm = TRUE),
    depth_at_OmegaCA_min = Depth[which.min(OmegaCAin)],
    depth_at_OmegaCA_max = Depth[which.max(OmegaCAin)],
    .groups = "drop"
  )
write_csv(rhizaria_summary, here::here("data/PRPOOS/rhizaria_summary.csv"))

### all
id_cols <- c("Station_ID", "Month_UTC", "Year_UTC")
taxa_cols <- setdiff(names(prpoos), id_cols)
taxa_cols <- taxa_cols[-c(1:7)]

prpoos_summary <- prpoos_merge %>%
  group_by(across(all_of(c(id_cols, taxa_cols)))) %>%
  summarise(
    pH_min = min(pHin, na.rm = TRUE),
    pH_max = max(pHin, na.rm = TRUE),
    pH_mean = mean(pHin, na.rm = TRUE),
    pH_median = median(pHin, na.rm = TRUE),
    pH_var = var(pHin),
    pH_range = max(pHin, na.rm = TRUE) - min(pHin, na.rm = TRUE),
    depth_at_pH_min = Depth[which.min(pHin)],
    depth_at_pH_max = Depth[which.max(pHin)],
    
    OmegaCA_min = min(OmegaCAin, na.rm = TRUE),
    OmegaCA_max = max(OmegaCAin, na.rm = TRUE),
    OmegaCA_mean = mean(OmegaCAin, na.rm = TRUE),
    OmegaCA_median = median(OmegaCAin, na.rm = TRUE),
    OmegaCA_var = var(OmegaCAin, na.rm = TRUE),
    OmegaCA_range = max(OmegaCAin, na.rm = TRUE) - min(OmegaCAin, na.rm = TRUE),
    depth_at_OmegaCA_min = Depth[which.min(OmegaCAin)],
    depth_at_OmegaCA_max = Depth[which.max(OmegaCAin)],
    
    CO3_min = min(CO3in, na.rm = TRUE),
    CO3_max = max(CO3in, na.rm = TRUE),
    CO3_mean = mean(CO3in, na.rm = TRUE),
    CO3_median = median(CO3in, na.rm = TRUE),
    CO3_var = var(CO3in, na.rm = TRUE),
    CO3_range = max(CO3in, na.rm = TRUE) - min(CO3in, na.rm = TRUE),
    depth_at_CO3_min = Depth[which.min(CO3in)],
    depth_at_CO3_max = Depth[which.max(CO3in)],
    .groups = "drop"
  )
write_csv(prpoos_summary, here::here("data/PRPOOS/prpoos_summary.csv"))
```

```{r}
# Visualization: Scatterplots for y (abundance) vs. x (pH)

### byrozoan_larvae
ggplot(byrozoan_larvae_summary) +
  aes(x = pH_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(byrozoan_larvae_summary$pH_mean, byrozoan_larvae_summary$Abundance)

### pteropoda_heteropoda
ggplot(pteropoda_heteropoda_summary) +
  aes(x = pH_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(pteropoda_heteropoda_summary$pH_mean, pteropoda_heteropoda_summary$Abundance)

### ostracods
ggplot(ostracods_summary) +
  aes(x = pH_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(ostracods_summary$pH_mean, ostracods_summary$Abundance)

### rhizaria
ggplot(rhizaria_summary) +
  aes(x = pH_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(rhizaria_summary$pH_mean, rhizaria_summary$Abundance)
```

```{r}
# Visualization: Scatterplots for y (abundance) vs. x (total calcium)

### byrozoan_larvae
ggplot(byrozoan_larvae_summary) +
  aes(x = OmegaCA_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(byrozoan_larvae_summary$OmegaCA_mean, byrozoan_larvae_summary$Abundance)

### pteropoda_heteropoda
ggplot(pteropoda_heteropoda_summary) +
  aes(x = OmegaCA_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(pteropoda_heteropoda_summary$OmegaCA_mean, pteropoda_heteropoda_summary$Abundance)

### ostracods
ggplot(ostracods_summary) +
  aes(x = OmegaCA_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(ostracods_summary$OmegaCA_mean, ostracods_summary$Abundance)

### rhizaria
ggplot(rhizaria_summary) +
  aes(x = OmegaCA_mean, y = log1p(Abundance)) +
  geom_point() +
  stat_cor(method = 'pearson')

cor(rhizaria_summary$OmegaCA_mean, rhizaria_summary$Abundance)
```

# Calcifiers EDA (effects of pH and anomolous abundance)
```{r}
prpoos_summary <- read_csv(here::here("data/PRPOOS/prpoos_summary.csv"))
prpoos_clean <- prpoos_summary %>%
  dplyr::select(-contains("Estimated C Biomass (mgC m-2)"))

names(prpoos_clean) <- str_replace(names(prpoos_clean), " \\(No\\. per m2\\)", "")

# Add seasonal component
prpoos_clean <- prpoos_clean %>%
  mutate(
    Season = case_when(
    Month_UTC %in% c(12, 1, 2) ~ "Winter",
      Month_UTC %in% c(3, 4, 5) ~ "Spring",
      Month_UTC %in% c(6, 7, 8) ~ "Summer",
      Month_UTC %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  relocate(Season, .after = Year_UTC)
```

```{r}
library(dplyr)
library(lubridate)
library(ggplot2)

# Create date variable
df_timeline <- prpoos_summary %>%
  mutate(
    YearMonth = ymd(paste(Year_UTC, Month_UTC, "01", sep = "-")),
    Station_ID = as.factor(Station_ID)  # ensure Station is treated as a category
  ) %>%
  dplyr::select(Station_ID, YearMonth) %>%
  distinct()  # remove duplicates if multiple observations per station/month

ggplot(df_timeline, aes(x = YearMonth, y = Station_ID)) +
  geom_point(alpha = 0.6, color = "darkblue") +
  labs(
    title = "Timeline of Observations by Station",
    x = "Date",
    y = "Station ID"
  ) +
  theme_minimal() +
  scale_x_date(date_labels = "%Y-%m", date_breaks = "6 months") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




station_counts <- prpoos_summary %>%
  group_by(Station_ID) %>%
  summarise(Record_Count = n()) %>%
  arrange(desc(Record_Count))


# Plot station record count
ggplot(station_counts, aes(x = reorder(Station_ID, -Record_Count), y = Record_Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Record Count per Station ID", x = "Station ID", y = "Number of Records") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

```{r}
calcifiers <- c("bryozoan_larvae Abundance",
                "pteropoda_heteropoda Abundance",
                "ostracods Abundance",
                "rhizaria Abundance")
abund_cols <- prpoos_clean %>% dplyr::select(contains("Abundance")) %>% names()
non_calcifiers <- setdiff(abund_cols, calcifiers)
```

```{r}
# Compute total abundance per group
prpoos_clean <- prpoos_clean %>%
  dplyr::mutate(
    calcifier_total = rowSums(dplyr::select(., all_of(calcifiers)), na.rm = TRUE),
    non_calcifier_total = rowSums(dplyr::select(., all_of(non_calcifiers)), na.rm = TRUE)
  )

# # Compute seasonal means
# seasonal_means <- prpoos_clean %>%
#   dplyr::group_by(Season) %>%
#   dplyr::summarise(
#     calcifier_mean = mean(calcifier_total, na.rm = TRUE),
#     non_calcifier_mean = mean(non_calcifier_total, na.rm = TRUE),
#     .groups = "drop"
#   )
# # Join means and compute anomalies
# prpoos_clean <- prpoos_clean %>%
#   left_join(seasonal_means, by = "Season") %>%
#   mutate(
#     calcifier_anomaly = calcifier_total - calcifier_mean,
#     non_calcifier_anomaly = non_calcifier_total - non_calcifier_mean
#   )
```

```{r}
# ggplot(prpoos_clean, aes(x = pH_mean, y = calcifier_anomaly, color = Season)) +
#   geom_point() +
#   geom_smooth(method = "lm") +
#   facet_wrap(~Season) +
#   labs(title = "Effect of pH on Calcifier Anomaly", x = "pH", y = "Anomaly") +
#   theme_minimal()
# 
# ggplot(prpoos_clean, aes(x = pH_mean, y = non_calcifier_anomaly, color = Season)) +
#   geom_point() +
#   geom_smooth(method = "lm") +
#   facet_wrap(~Season) +
#   labs(title = "Effect of pH on Non-Calcifier Anomaly", x = "pH", y = "Anomaly") +
#   theme_minimal()

ggplot(prpoos_clean, aes(x = pH_mean, y = calcifier_total, color = Season)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Season) +
  labs(
    title = "Calcifier Abundance vs. avg pH by Season",
    x = "Average pH",
    y = "Total Abundance (no. per m2)"
  ) +
  theme_minimal()

ggplot(prpoos_clean, aes(x = pH_mean, y = non_calcifier_total, color = Season)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Season) +
  labs(
    title = "Non-Calcifier Abundance vs. avg pH by Season",
    x = "Average pH",
    y = "Total Abundance (no. per m2)"
  ) +
  theme_minimal()

ggplot(prpoos_clean, aes(x = OmegaCA_mean, y = calcifier_total, color = Season)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Season) +
  labs(
    title = "Calcifier Abundance vs. Total Calcium by Season",
    x = "OmegaCA",
    y = "Total Abundance"
  ) +
  theme_minimal()

ggplot(prpoos_clean, aes(x = OmegaCA_mean, y = non_calcifier_total, color = Season)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE) +
  facet_wrap(~Season) +
  labs(
    title = "Non-Calcifier Abundance vs. Total Calcium by Season",
    x = "OmegaCA",
    y = "Total Abundance"
  ) +
  theme_minimal()

```

# ZooDB Calcifiers
## data/ZooDB
**Mollusca** (4 taxa) - calcite or aragonite:
- Mollusca Euthecosomata (mollusca1)
- Mollusca Gymnosaomata (mollusca2)
- Mollusca Heteropoda Atlantidae (mollusca3)
- Mollusca Pseudothecosomata (mollusca4)
**Ostracoda** - shell is complex arrangement of low-Mg calcite crystals and organic components (ostracoda)
**Radiolaria** - non-calcifying (Silicate) (radiolaria)
**Foraminifera** - calcite or aragonite (foraminifera)
- (I'm assuming she meant foraminifera not formaminefera)

* Next steps: merge with carb chem data

```{r}
# Read in datasets
mollusca1 <- read_csv(here::here("data/ZooDB/mollusca_euthecosomata.txt"), skip = 5)
mollusca2 <- read_csv(here::here("data/ZooDB/mollusca_gymnosaomata.txt"), skip = 5)
mollusca3 <- read_csv(here::here("data/ZooDB/mollusca_heteropoda_atlantidae.txt"), skip = 5)
mollusca4 <- read_csv(here::here("data/ZooDB/mollusca_pseudothecosomata.txt"), skip = 5)
ostracoda <- read_csv(here::here("data/ZooDB/ostracoda.txt"), skip = 5)
radiolaria <- read_csv(here::here("data/ZooDB/radiolaria.txt"), skip = 5)
foraminifera <- read_csv(here::here("data/ZooDB/foraminifera.txt"), skip = 5)
```

```{r}
# Preprocessing datasets
## Combine `Line` and `Station` (in that order) to create `Station_ID` variable
## Extract Month and Year from `Station date` into separate variables

### mollusca1
mollusca1$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(mollusca1$Line)),
  sprintf('%05.1f', as.numeric(mollusca1$Station)),
  sep = ' '
)
mollusca1 <- mollusca1 %>%
  relocate(Station_ID, .after = Station)
mollusca1 <- mollusca1 %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### mollusca2
mollusca2$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(mollusca2$Line)),
  sprintf('%05.1f', as.numeric(mollusca2$Station)),
  sep = ' '
)
mollusca2 <- mollusca2 %>%
  relocate(Station_ID, .after = Station)
mollusca2 <- mollusca2 %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### mollusca3
mollusca3$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(mollusca3$Line)),
  sprintf('%05.1f', as.numeric(mollusca3$Station)),
  sep = ' '
)
mollusca3 <- mollusca3 %>%
  relocate(Station_ID, .after = Station)
mollusca3 <- mollusca3 %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### mollusca4
mollusca4$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(mollusca4$Line)),
  sprintf('%05.1f', as.numeric(mollusca4$Station)),
  sep = ' '
)
mollusca4 <- mollusca4 %>%
  relocate(Station_ID, .after = Station)
mollusca4 <- mollusca4 %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### ostracoda
ostracoda$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(ostracoda$Line)),
  sprintf('%05.1f', as.numeric(ostracoda$Station)),
  sep = ' '
)
ostracoda <- ostracoda %>%
  relocate(Station_ID, .after = Station)
ostracoda <- ostracoda %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### radiolaria
radiolaria$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(radiolaria$Line)),
  sprintf('%05.1f', as.numeric(radiolaria$Station)),
  sep = ' '
)
radiolaria <- radiolaria %>%
  relocate(Station_ID, .after = Station)
radiolaria <- radiolaria %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )

### foraminifera
foraminifera$Station_ID <- paste(
  sprintf('%05.1f', as.numeric(foraminifera$Line)),
  sprintf('%05.1f', as.numeric(foraminifera$Station)),
  sep = ' '
)
foraminifera <- foraminifera %>%
  relocate(Station_ID, .after = Station)
foraminifera <- foraminifera %>%
  mutate(
    Year_UTC = year(Date),
    Month_UTC = month(Date),
    Day_UTC = day(Date),
    .after = Date
  )
```

```{r}
# Merge

### mollusca1
mollusca1_merge <- co2sys %>%
  left_join(mollusca1, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### mollusca2
mollusca2_merge <- co2sys %>%
  left_join(mollusca2, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### mollusca3
mollusca3_merge <- co2sys %>%
  left_join(mollusca3, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### mollusca4
mollusca4_merge <- co2sys %>%
  left_join(mollusca4, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### ostracoda
ostracoda_merge <- co2sys %>%
  left_join(ostracoda, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### radiolaria
radiolaria_merge <- co2sys %>%
  left_join(radiolaria, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

### foraminifera
foraminifera_merge <- co2sys %>%
  left_join(foraminifera, by = c("Station_ID", "Month_UTC", "Year_UTC")) %>%
  na.omit()

length(intersect(mollusca1$Station_ID, co2sys$Station_ID))
length(intersect(mollusca2$Station_ID, co2sys$Station_ID))
length(intersect(mollusca3$Station_ID, co2sys$Station_ID))
length(intersect(mollusca4$Station_ID, co2sys$Station_ID))
length(intersect(ostracoda$Station_ID, co2sys$Station_ID))
length(intersect(radiolaria$Station_ID, co2sys$Station_ID))
length(intersect(foraminifera$Station_ID, co2sys$Station_ID))
```


