---
title: "eda-merged-bottle-data"
author: "Liuqian Bao"
format: html
editor: visual
---

# Spacial and temporal information

```{r}
# Load necessary libraries
library(ggplot2)
library(readr)
# read in merged bottle data
merged_bottle_data <- read_csv("../data/merged_bottle_data.csv")
# First plot: Date vs Station ID
ggplot(merged_bottle_data, aes(x = Date, y = Station_ID)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "Visualization of Date vs Station ID", x = "Date", y = "Station ID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
library(mapview)  # For interactive maps
library(sf)       # For spatial data handling
library(dplyr)    # For data manipulation

# Count the number of observations per Station_ID
station_counts <- merged_bottle_data %>%
  group_by(Station_ID) %>%
  summarise(Observation_Count = n())  # Count occurrences per Station_ID

# Merge the count data back to the original dataset
merged_bottle_ct <- merged_bottle_data %>%
  left_join(station_counts, by = "Station_ID")

# Convert to spatial format using sf (assuming 'Latitude' and 'Longitude' columns exist)
merged_bottle_sf <- st_as_sf(merged_bottle_ct, 
                             coords = c("Longitude", "Latitude"), 
                             crs = 4326)  # WGS 84 (standard lat/lon system)

# Create an interactive map
mapview(merged_bottle_sf, 
        zcol = "Observation_Count",  # Color by number of observations
        layer.name = "Stations (Observations Count)", 
        legend = TRUE)
```

# Variable descriptions

## Sample identifiers

`Cst_Cnt`: Cast Count - All CalCOFI casts ever conducted, consecutively numbered

`Btl_Cnt`: Bottle Count - All CalCOFI bottles ever sampled, consecutively numbered

`Station_ID`: Line and Station \[Line\] \[Station\]

## Spacial/temporal info

`Date`: date object (yyyy-mm-dd)

`Depthm`: Bottle depth in meters

`Latitude`:

`Longitude`:

## Measurements

### from calcofi hydro bottle data:

`T_degC`: Water temperature in degrees Celsius (degrees Celsius)

`Salnty`: Salinity (Practical Salinity Scale 1978)

`O2ml_L`: Milliliters oxygen per liter of seawater

`STheta`: Potential Density (Sigma Theta) (kilograms per cubic meter)

`O2Sat`: Oxygen percent saturation

`Oxy_ $\mu$ mol/Kg`: Oxygen micromoles per kilogram seawater

`ChlorA`: Migrograms Chlorophyll-a per liter seawater, measured fluorometrically

`Phaeop`: Micrograms Phaeopigment per liter seawater, measured fluormetrically

`PO4uM`: Micromoles Phosphate per liter of seawater

`SiO3uM`: Micromoles Silicate per liter of seawater

`NO2uM`: Micromoles Nitrite per liter of seawater

`NO3uM`: Micromoles Nitrate per liter of seawater

`NH3uM`: Micromoles Ammonia per liter of seawater

`DIC1`: Dissolved Inorganic Carbon micromoles per kilogram solution

`DIC2`: Dissolved Inorganic Carbon micromoles per kilogram solution (on a replicate sample)

`TA1`: Total Alkalinity micromoles per kilogram solution

`TA2`: Total Alkalinity micromoles per kilogram solution (on a replicate sample)

`pH1`: pH (the degree of acidity/alkalinity of a solution)

`pH2`: pH (the degree of acidity/alkalinity of a solution) on a replicate sample

(Obtained from: <https://calcofi.org/data/oceanographic-data/bottle-database/>)

### from carbonate chemistry data:

| **Dissolved Inorganic Carbon** |
|:------------------------------:|
|       **Abbreviation:**        |
|           **Unit:**            |

| **Total alkalinity** |
|:--------------------:|
|  **Abbreviation:**   |
|      **Unit:**       |

| **Sea water practical salinity** |
|:--------------------------------:|
|        **Abbreviation:**         |
|            **Unit:**             |

| **Sea water temperature** |
|:-------------------------:|
|     **Abbreviation:**     |
|         **Unit:**         |

(Citation: Keeling, Charles D.; Lueker, Timothy J.; Emanuele, Guy; Dickson, Andrew G.; Martz, Todd R.; Wolfe, Wiley H.; Mau, Aaron (2025). Discrete profile dissolved inorganic carbon, total alkalinity, water temperature and salinity measurements collected and assembled from a number of cruises on many different ships for the California Cooperative Oceanic Fisheries Investigations program (CalCOFI) in the North East Pacific Ocean Coastal area from 1983-03-19 to 2021-07-20 (NCEI Accession 0301029). NOAA National Centers for Environmental Information. Unpublished Dataset. https://www.ncei.noaa.gov/archive/accession/0301029. Accessed Jan 2nd 2025.)
