---
title: "2nd part merge(Zoo+Carbonate)"
format: html
editor: visual
---
<<<<<<< Updated upstream

```{r}
# Load libraries
library(readr)
library(dplyr)

# Load data
oah_bottle <- read_csv("~/Desktop/Pstat197BC/capstone-calcofi-seagrant/data/carbonate_chem_bottle.csv")
zooplankton <- read_csv("~/Desktop/Pstat197BC/capstone-calcofi-seagrant/data/zooplankton.csv")


# `Year_UTC`, `Month_UTC`, `Day_UTC` in oah_bottle
oah_bottle <- oah_bottle %>%
  filter(!is.na(Year_UTC)) %>%  # romove NaN lines
  mutate(
    Date = sprintf("%02d/%02d/%04d", Month_UTC, Day_UTC, Year_UTC),  #MM/DD/YYYY
    Station_ID = trimws(as.character(Station_ID))  # process Station_ID
  )

# `Tow_Date` and `Sta_ID` in zooplankton
zooplankton <- zooplankton %>%
  rename(Station_ID = Sta_ID) %>%
  mutate(
    Station_ID = trimws(as.character(Station_ID)),
    Tow_Date = as.character(Tow_Date)  # string check
  )

# Merge
merged_zoop_data <- inner_join(
  oah_bottle, 
  zooplankton,
  by = join_by(Date == Tow_Date, Station_ID)
)


#write_csv(merged_zoop_data, "merged_zoop_data.csv")


nrow(merged_zoop_data)
# We have 456 lines for the merge data

```

```{r}
library(ggplot2)

merged_zoop_data <- read_csv("merged_zoop_data.csv")

# Ensure Date column is in Date format
merged_zoop_data <- merged_zoop_data %>%
  mutate(Date = as.Date(Date, format="%m/%d/%Y"))

# Compute records per station
station_counts <- merged_zoop_data %>%
  group_by(Station_ID) %>%
  summarise(Record_Count = n()) %>%
  arrange(desc(Record_Count))

# Compute records per date
date_counts <- merged_zoop_data %>%
  group_by(Date) %>%
  summarise(Record_Count = n()) %>%
  arrange(Date)

# Plot station record count
ggplot(station_counts, aes(x = reorder(Station_ID, -Record_Count), y = Record_Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Record Count per Station ID", x = "Station ID", y = "Number of Records") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Plot date record count
ggplot(date_counts, aes(x = Date, y = Record_Count)) +
  geom_line(color = "darkred") +
  geom_point(size = 2, color = "darkred") +
  labs(title = "Record Count Over Time", x = "Date", y = "Number of Records") +
  theme_minimal()

```
Station Data Distribution (First Plot):
1. Station ID 090.0 090.0 has the highest number of records, exceeding 120.
2. Other stations have fewer records, and some stations have sparse data coverage.

Temporal Data Distribution (Second Plot):
1. Most data is concentrated between 1983-1992 and 2012-2017.
2. There is a gap between 1995-2003, indicating that no sampling was conducted or records were missing during this period.

## Scatter plot: Relationship Between Ocean Chemistry and Zooplankton Volume

```{r}
ggplot(merged_zoop_data, aes(x = DIC, y = Ttl_PVolC3)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "DIC vs. Total Plankton Volume",
       x = "Dissolved Inorganic Carbon (DIC, µmol/kg)", 
       y = "Total Plankton Volume (cm³)") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "DIC_vs_Plankton.png"))
```
## Scatter plot: Temperature vs. Total Plankton Volume
```{r}
# Scatter plot: Temperature vs. Total Plankton Volume
ggplot(merged_zoop_data, aes(x = CTDTEMP_ITS90, y = Ttl_PVolC3)) +
  geom_point(alpha = 0.6, color = "red") +
  labs(title = "Temperature vs. Total Plankton Volume",
       x = "Temperature (°C)", 
       y = "Total Plankton Volume (cm³)") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Temp_vs_Plankton.png"))
```
```{r}
library(ggplot2)
library(reshape2)
#library(ggcorrplot)

# Compute correlation
cor_matrix <- cor(merged_zoop_data[, c("CTDTEMP_ITS90", "Ttl_PVolC3", "DIC", "Salinity_PSS78", "TA")], use = "complete.obs")

# Convert to long format
melted_cor <- melt(cor_matrix)

# Plot with ggplot2
ggplot(melted_cor, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red") +
  labs(title = "Correlation Matrix: Ocean Chemistry vs Plankton") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
        axis.text.y = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, face = "bold"))



```
```{r}
library(corrplot)
corrplot(cor_matrix, method = "color", col = colorRampPalette(c("blue", "white", "red"))(200),
         addCoef.col = "black", number.cex = 0.7, tl.cex = 1, tl.srt = 45, title = "Correlation Matrix")
ggsave(file.path("JiajiaF merged plots", "Correlation Matrix.png"))
```

## Spatial Analysis: Distribution of Plankton & Chemistry

```{r}
ggplot(merged_zoop_data, aes(x = Longitude, y = Latitude, color = DIC)) +
  geom_point(size = 3, alpha = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Station Locations (Colored by DIC)",
       x = "Longitude", y = "Latitude", color = "DIC (µmol/kg)") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Station_Locations_DIC.png"))
```
## 3. Seasonal Patterns
```{r}
# Time Series: DIC, Temperature, Salinity, Plankton Volume over Time
ggplot(merged_zoop_data, aes(x = Date)) +
  geom_line(aes(y = DIC, color = "DIC")) +
  geom_line(aes(y = Salinity_PSS78 * 100, color = "Salinity (scaled)")) +
  geom_line(aes(y = CTDTEMP_ITS90 * 10, color = "Temperature (scaled)")) +
  geom_line(aes(y = Ttl_PVolC3, color = "Plankton Volume")) +
  labs(title = "Time Series of Ocean Chemistry & Plankton Volume",
       x = "Date", y = "Scaled Values") +
  scale_color_manual(values = c("DIC" = "blue", "Salinity (scaled)" = "purple", 
                                "Temperature (scaled)" = "red", "Plankton Volume" = "green")) +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Time_Series_Chemistry_Plankton.png"))
```

## 4. Depth-Dependent Changes
```{r}
# Scatter Plot: Depth vs. DIC
ggplot(merged_zoop_data, aes(x = Depth, y = DIC)) +
  geom_point(alpha = 0.6, color = "blue") +
  labs(title = "Depth vs. DIC",
       x = "Depth (m)", y = "DIC (µmol/kg)") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Depth_vs_DIC.png"))

# Scatter Plot: Depth vs. Plankton Volume
ggplot(merged_zoop_data, aes(x = Depth, y = Ttl_PVolC3)) +
  geom_point(alpha = 0.6, color = "green") +
  labs(title = "Depth vs. Total Plankton Volume",
       x = "Depth (m)", y = "Total Plankton Volume (cm³)") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Depth_vs_Plankton.png"))
```
## 5. Number of Records Per Station & Date
```{r}
# Compute records per station
station_counts <- merged_zoop_data %>%
  group_by(Station_ID) %>%
  summarise(Record_Count = n()) %>%
  arrange(desc(Record_Count))

# Plot station record count
ggplot(station_counts, aes(x = reorder(Station_ID, -Record_Count), y = Record_Count)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  labs(title = "Record Count per Station ID", x = "Station ID", y = "Number of Records") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
ggsave(file.path("JiajiaF merged plots", "Station_Record_Count.png"))

# Compute records per date
date_counts <- merged_zoop_data %>%
  group_by(Date) %>%
  summarise(Record_Count = n()) %>%
  arrange(Date)

# Plot date record count
ggplot(date_counts, aes(x = Date, y = Record_Count)) +
  geom_line(color = "darkred") +
  geom_point(size = 2, color = "darkred") +
  labs(title = "Record Count Over Time", x = "Date", y = "Number of Records") +
  theme_minimal()
ggsave(file.path("JiajiaF merged plots", "Date_Record_Count.png"))
```

