```{r}
library(ggplot2)
library(sf)
library(rnaturalearth)
```


# Preprocessing
```{r}
# load new zooplankton data
zoop_data <- read.csv('../data/zoop_data/Zooplankton-new.csv')

# load carbonate chem bottle data
cc_data <- read.csv('../data/carbonate_chem_data/carbonate_chem_bottle.csv')[-1,]
```

```{r}
head(zoop_data)
```

```{r}
head(cc_data)
```


```{r}
# format Station_ID column
zoop_data$Station_ID <- paste(
  sprintf('%05.1f', zoop_data$line),
  sprintf('%05.1f', zoop_data$station),
  sep = ' '
)
```


```{r}
# format date in carbonate chem data
cc_data <- cc_data %>%
  mutate(
    Date = as.Date(
      paste(Month_UTC, Day_UTC, Year_UTC, sep = "/"),
      tryFormats = c("%m/%d/%Y")
    ),
    .before = Year_UTC
  ) %>%
  mutate(
    Depth = as.double(Depth)
  )

# format date in zooplankton data
zoop_data$Date = as.Date(zoop_data$time)
```


# Perform Join
```{r}
# join data on matching station ID and date
merged_cc_zoop_data <- inner_join(
  cc_data,
  zoop_data,
  by = join_by(Station_ID, Date)
)

head(merged_cc_zoop_data)
```


```{r}
unique(merged_cc_zoop_data$Date)
```

```{r}
unique(merged_cc_zoop_data$Station_ID)
```






# Updated Merge: Average by month
```{r}
zoop_data$Month_UTC <- as.integer(format(zoop_data$Date, '%m'))
zoop_data$Year_UTC <- as.integer(format(zoop_data$Date, '%Y'))
```

```{r}
head(zoop_data)
```


```{r}
# numeric columns to be averaged
zoop_data_num_cols <- names(select_if(zoop_data, is.numeric))

# group zoop data by month
zoop_data_by_month <- zoop_data[, c('Station_ID', zoop_data_num_cols)] %>%
  group_by(Station_ID, Year_UTC, Month_UTC) %>%
  summarise(across(everything(), mean), .groups = 'drop')  %>%
  as.data.frame()
```


```{r}
# numeric columns to be averaged
cc_data_num_cols <- names(select_if(cc_data, is.numeric))

# group cc data by month
cc_data_by_month <- cc_data[, c('Station_ID', cc_data_num_cols)] %>%
  group_by(Station_ID, Year_UTC, Month_UTC) %>%
  summarise(across(everything(), mean), .groups = 'drop') %>%
  as.data.frame()
```


# Perform Join
```{r}
merged_cc_zoop_data <- inner_join(
  cc_data_by_month,
  zoop_data_by_month,
  by = join_by(Station_ID, Year_UTC, Month_UTC)
)
```

```{r}
merged_cc_zoop_data[with(merged_cc_zoop_data, order(Year_UTC, Month_UTC)),]
```


# EDA on New Merged Dataset
```{r}
# number of unique stations
length(unique(merged_cc_zoop_data$Station_ID))
```

```{r}
# number of unique months
count(unique(merged_cc_zoop_data[, c('Year_UTC', 'Month_UTC')]))
```
50 unique stations, 114 different months


```{r}
# number of observations by station
count_by_station <- merged_cc_zoop_data %>% 
  count(Station_ID, sort=TRUE) %>%
  rename(Count=n)

# bar plot
bar_count_by_station <- ggplot(data=count_by_station, aes(x=Count, y=reorder(Station_ID, Count))) +
  geom_bar(stat='identity', fill='steelblue') + 
  ylab('Station_ID') +
  ggtitle('Observation Count by Station') +
  theme_minimal()

# save figure
ggsave(path='../images/merged_zoop_EDA', filename='obs_count_by_station.png', width=9, height=6)

bar_count_by_station
```

```{r}
# create date column
merged_cc_zoop_data$Date <- as.Date(
  with(merged_cc_zoop_data, sprintf('%d-%02d-01', Year_UTC, Month_UTC))
)
```
 

```{r}
# number of observations by date
count_by_date <- merged_cc_zoop_data %>%
  count(Date, sort=TRUE) %>%
  rename(Count=n)

# bar plot
bar_count_by_date <- ggplot(data=count_by_date, aes(x=Date, y=Count)) +
  geom_bar(stat='identity', fill='steelblue', width = 80) + 
  ggtitle('Observation Count by Date') +
  theme_minimal()
  
# save figure
ggsave(path='../images/merged_zoop_EDA', filename='obs_count_by_date.png', width=9, height=6)
  
bar_count_by_date
```



```{r}
# filter coordinate columns and count observations by station
coords_merged_zoop <- merged_cc_zoop_data[, c('Station_ID', 'latitude', 'longitude')] %>%
  count(Station_ID, latitude, longitude) %>%
  rename(Num_Observations = n)

# find min and max coordinates
min_lon <- min(coords_merged_zoop[, 'longitude'])
max_lon <- max(coords_merged_zoop[, 'longitude'])

min_lat <- min(coords_merged_zoop[, 'latitude'])
max_lat <- max(coords_merged_zoop[, 'latitude'])
```

```{r}
# plot observed station location with density of observations
world <- ne_countries(scale = "medium", returnclass = "sf")

station_loc_map_new <- ggplot(data=world) + 
  geom_sf(fill = "antiquewhite1") + 
  geom_point(data=coords_merged_zoop, aes(x=longitude, y=latitude, size=Num_Observations)) + 
  coord_sf(xlim = c(min_lon - 1, max_lon + 4), ylim = c(min_lat - 1, max_lat + 1), expand=FALSE) + 
  theme(panel.grid.major = element_line(color = gray(0.5), linetype = "solid", 
        linewidth = 0.5), panel.background = element_rect(fill = "aliceblue")) + 
  ggtitle('Station Locations of Updated Merged Zooplankton Data')

# save figure
ggsave(path='../images/merged_zoop_EDA', filename='station_loc_map_new.png')

station_loc_map_new
```

# Distribution of Observations Across Time


```{r}
# plot each observation by time and station
timeline_obs_station_new <- ggplot(data=merged_cc_zoop_data, aes(x=Date, y=Station_ID)) +
  geom_point(size=1, shape=1) + 
  scale_x_date(name = '', date_breaks = '5 years', expand = expansion(mult = c(0.05, 0.1))) + 
  theme_minimal() + 
  ggtitle('Timeline of Observations by Station')

timeline_obs_station_new
```







# Salinity and Depth
```{r}
salinity_depth_scatter <- ggplot(data=merged_cc_zoop_data, aes(x=Salinity_PSS78, y=Depth)) + 
  geom_point(size=1, shape=1) + 
  scale_y_reverse() + 
  ggtitle('Scatterplot of Salinity and Depth') +
  theme_minimal()

# save figure
ggsave(path='../images/merged_zoop_EDA', filename='salinity_depth_scatter.png', width=9, height=6)

salinity_depth_scatter
```


# Zooplankton and Depth
```{r}
total_plankton_depth_scatter <- ggplot(data=merged_cc_zoop_data, aes(x=total_plankton, y=Depth)) + 
  geom_point(size=1, shape=1) + 
  scale_y_reverse() + 
  ggtitle('Scatterplot of Total Plankton and Depth') +
  theme_minimal()

# save figure
ggsave(path='../images/merged_zoop_EDA', filename='total_plankton_depth_scatter.png', width=9, height=6)

zooplankton_depth_scatter
```

```{r}
small_plankton_depth_scatter <- ggplot(data=merged_cc_zoop_data, aes(x=small_plankton, y=Depth)) + 
  geom_point(size=1, shape=1) + 
  scale_y_reverse() + 
  ggtitle('Scatterplot of Small Plankton and Depth') + 
  theme_minimal()

# save figure
ggsave(path='../images/merged_zoop_EDA', filename='small_plankton_depth_scatter.png', width=9, height=6)

small_plankton_depth_scatter
```

# Outliers
Each scatterplot has a group of 4 observations that are measured at a much greater depth
```{r}

```


# Salinity and Plankton
```{r}
ggplot(data=merged_cc_zoop_data, aes(x=Salinity_PSS78, y=total_plankton)) + 
  geom_bin_2d(bins=80) + 
  theme_minimal()
```



```{r}
unique(merged_cc_zoop_data$Station_ID)
```

# Time Series of Plankton Data



