---
title: "Untitled"
format: html
editor: visual
---

## Exploring the Impacts of Ocean Acidification

### 1. Introduction

Objectives:

-   Explore the temporal and seasonal variability in zooplankton abundance.
-   Analyze correlations between key ocean acidification variables (pH, CO₃²⁻, ΩCA) and zooplankton abundance.
-   Model the effects of these environmental factors on the abundance of selected species.

### 2. Methods

#### 2.1 Data Preprocessing

```{r}
library(tidyverse)
library(readr)
library(GGally)
library(broom)
library(car)
library(mgcv)
```

```{r}
df <- read_csv("/Users/fengci/Desktop/Pstat197BC/capstone-calcofi-seagrant/data/PRPOOS/prpoos_summary_namecleaned.csv")
```

Choose spoecies based on Ecological Roles:

-   **Filter-feeding zooplankton (**highly sensitive to water chemistry changes): *Salps*.
-   **Predators / Higher trophic level (**feed on other zooplankton): *Chaetognatha*.
-   **Crustacean zooplankton (**large population sizes and commonly used in monitoring): *Euphausiids*.
-   **Calcifying soft-bodied organisms (**have aragonite shells and directly affected by pH and OmegaCA): *Pteropoda heteropoda.*

Predictors:

pH_mean, OmegaCA_mean, CO3_mean, and depth_at_pH_min

```{r}
# Filter the columns required for analysis
df_selected <- df %>%
  select(
    Year_UTC, Month_UTC, depth_at_pH_min, pH_mean, OmegaCA_mean, CO3_mean,
    salpsAbundance, chaetognathaAbundance, euphausiidsAbundance, pteropoda_heteropodaAbundance
  ) %>%
  drop_na()

df_selected
```

#### 2.2. Exploratory Data Analysis (EDA)

Interannual Trends in Zooplankton Abundance
```{r}
df_selected %>%
  pivot_longer(cols = ends_with("Abundance"),
               names_to = "Species", values_to = "Abundance") %>%
  ggplot(aes(x = Year_UTC, y = Abundance, color = Species)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE) +
  facet_wrap(~Species, scales = "free_y") +
  theme_minimal() +
  labs(title = "Zooplankton Abundance Over Years", y = "Abundance")
```
Monthly Variation in Abundance
```{r}
df_selected %>%
  pivot_longer(cols = ends_with("Abundance"),
               names_to = "Species", values_to = "Abundance") %>%
  ggplot(aes(x = as.factor(Month_UTC), y = Abundance, fill = Species)) +
  geom_boxplot() +
  facet_wrap(~Species, scales = "free_y") +
  labs(title = "Abundance by Month", x = "Month") +
  theme_minimal()
```


Pairwise Correlation Between Species and Environmental Variables

```{r}
ggpairs(df_selected, columns = c("pH_mean", "OmegaCA_mean", "CO3_mean",
                                 "salpsAbundance", "chaetognathaAbundance",
                                 "euphausiidsAbundance", "pteropoda_heteropodaAbundance"))
```

#### 2.3. Regression Modeling
Linear Regression (Example: Euphausiids)
```{r}

model_eup <- lm(euphausiidsAbundance ~ pH_mean + OmegaCA_mean + CO3_mean + depth_at_pH_min,
                data = df_selected)
summary(model_eup)

```

Multicollinearity Check (VIF)
```{r}
vif(model_eup)
```

Residual Diagnostics
```{r}
augment(model_eup) %>%
  ggplot(aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  theme_minimal() +
  labs(title = "Residuals vs Fitted", x = "Fitted values", y = "Residuals")
```

Try GAM if non-linear pattern is detected
```{r}
gam_model <- gam(euphausiidsAbundance ~ s(OmegaCA_mean) + s(CO3_mean) + s(depth_at_pH_min),
                 data = df_selected)
summary(gam_model)
```

```{r}
# ========== Helper Function ==========
analyze_species <- function(species_name) {
  formula <- as.formula(paste0(species_name, " ~ pH_mean + OmegaCA_mean + CO3_mean + depth_at_pH_min"))
  model <- lm(formula, data = df_selected)
  gam_formula <- as.formula(paste0(species_name, " ~ s(OmegaCA_mean) + s(CO3_mean) + s(depth_at_pH_min)"))
  gam_mod <- gam(gam_formula, data = df_selected)

  cat("\n==============================\n")
  cat("LINEAR MODEL for:", species_name, "\n")
  print(summary(model))
  cat("\n--- VIF ---\n")
  print(vif(model))

  # Residual Plot
  print(
    augment(model) %>%
      ggplot(aes(x = .fitted, y = .resid)) +
      geom_point(alpha = 0.6) +
      geom_hline(yintercept = 0, linetype = "dashed") +
      theme_minimal() +
      labs(title = paste("Residuals vs Fitted:", species_name),
           x = "Fitted values", y = "Residuals")
  )

  # Optional GAM summary
  cat("\n--- GAM Model Summary ---\n")
  print(summary(gam_mod))
}

# ========== Run for all 4 species ==========
species_list <- c("salpsAbundance", "chaetognathaAbundance",
                  "euphausiidsAbundance", "pteropoda_heteropodaAbundance")

lapply(species_list, analyze_species)

```

