---
title: "zoop_eda_liuqian"
author: "Liuqian Bao"
date: "2025-02-12"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Variable descriptions: `small_plankton`: Standardized volume of plankton with individual displacement volumes <5 ml(the eggs and larvae of fish found mainly in the upper 200 meters of the water column, also called the near-surface waters.)
`total plankton`: Standardized volume of plankton in the sample
`volume_sampled`: Volume of water sampled by the net

For net tow samples, laboratory processing begins with plankton volume determination of the bongo net sample that is preserved in formalin. Two volumes are measured by displacement for each sample: total volume and "small volume". Total volume includes everything in the sample except any non-planktonic organisms such as juvenile and adult fish, large cephalopods, and pelagic crabs.

The total volume is determined by measuring the volume of plankton and formalin together, and subtracting the volume of formalin remaining once the plankton has been strained out. "Small volume" is determined using the same method, however, small volume is the total volume with the large plankters removed (salps and jellies).

```{r}
# Load necessary libraries
library(ggplot2)
library(readr)
# read in merged bottle data
merged_zoop_avg_monthly <- read_csv("../data/merged_zoop_avg_monthly.csv")
# First plot: Date vs Station ID
ggplot(merged_zoop_avg_monthly, aes(x = Date.x, y = Station_ID)) +
  geom_point(alpha = 0.5, size = 1) +
  labs(title = "Visualization of Date vs Station ID", x = "Date.x", y = "Station ID") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
merged_zoop_avg_monthly <- merged_zoop_avg_monthly %>%
  mutate(YearMonth = as.Date(paste(Year_UTC, Month_UTC, "01", sep = "-"), 
                             format = "%Y-%m-%d"))
ggplot(merged_zoop_avg_monthly, aes(x = YearMonth, y = Station_ID, size = total_plankton, color = total_plankton)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c() +          
  scale_size(range = c(1, 5)) +     # control the size range of the points
  labs(title = "Scatter Plot with Size and Color Representing total_plankton",
       x = "Year-Month",
       y = "Station",
       color = "total_plankton",
       size = "total_plankton") +
  theme_minimal()
```

```{r}
# time series for one station 
library(dplyr)
zmerged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly %>%
  filter(Station_ID == "083.3 042.0")
ggplot(zmerged_zoop_avg_monthly_filtered, aes(x = YearMonth, y = total_plankton)) +
  geom_line() +
  labs(
    title = "Time Series for station 083.3 042.0",
    x = "Year-Month",
    y = "total_plankton"
  ) +
  theme_minimal()
```

```{r}
# time series for one station 
library(dplyr)
zmerged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly %>%
  filter(Station_ID == "086.7 035.0")
ggplot(zmerged_zoop_avg_monthly_filtered, aes(x = YearMonth, y = total_plankton)) +
  geom_line() +
  labs(
    title = "Time Series for station 086.7 035.0",
    x = "Year-Month",
    y = "total_plankton"
  ) +
  theme_minimal()
```

```{r}
# time series for one station 
library(dplyr)
zmerged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly %>%
  filter(Station_ID == "093.3 030.0")
ggplot(zmerged_zoop_avg_monthly_filtered, aes(x = YearMonth, y = total_plankton)) +
  geom_line() +
  labs(
    title = "Time Series for station 093.3 030.0",
    x = "Year-Month",
    y = "total_plankton"
  ) +
  theme_minimal()
```

```{r}
library(dplyr)
library(ggplot2)
library(lubridate)

# Get all unique station IDs and sort them
all_stations <- sort(unique(merged_zoop_avg_monthly_filtered$Station_ID))

# Compute indices for evenly spaced selection of 16 stations
selected_indices <- round(seq(1, length(all_stations), length.out = 16))
stations_to_plot <- all_stations[selected_indices]

# Filter the dataset for these 16 stations
subset_data <- merged_zoop_avg_monthly_filtered %>%
  filter(Station_ID %in% stations_to_plot)

# Create the faceted time series plot
ggplot(subset_data, aes(x = YearMonth, y = total_plankton)) +
  geom_line() +
  facet_wrap(~ Station_ID, ncol = 4) +  # Adjust ncol as desired
  labs(
    title = "Time Series for 16 'Evenly Spaced' Stations",
    x = "Year-Month",
    y = "Total Plankton"
  ) +
  scale_x_date(date_breaks = "4 year", date_labels = "%Y") +
  theme_minimal()

```


```{r}
# remove outliers
merged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly %>%
  filter(total_plankton <= 2000, Year_UTC >= 2005) %>% 
  #select(CTDTEMP_ITS90, Salinity_PSS78, DIC, TA, total_plankton, small_plankton, large_plankton) %>%
  drop_na()

ggplot(zmerged_zoop_avg_monthly_filtered, aes(x = YearMonth, y = Station_ID, size = total_plankton, color = total_plankton)) +
  geom_point(alpha = 0.7) +
  scale_color_viridis_c() +          
  scale_size(range = c(2, 6)) +     # control the size range of the points
  labs(title = "Scatter Plot with Size and Color Representing total_plankton",
       x = "Year-Month",
       y = "Station",
       color = "total_plankton",
       size = "total_plankton") +
  theme_minimal()
```

```{r}
library(naniar)
vis_miss(merged_zoop_avg_monthly) 
```

```{r}
library(ggcorrplot)
merged_zoop_avg_monthly_filtered %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value)) +
  geom_histogram(bins = 30, fill = "blue", alpha = 0.5) +
  facet_wrap(~Variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Environmental Factors and Plankton Counts")
       

```

```{r}
library(corrplot)

# Compute correlation matrix
cor_matrix <- cor(merged_zoop_avg_monthly_filtered, use = "complete.obs")

# Visualize the correlation matrix
corrplot(cor_matrix, method = "color", type = "upper", tl.cex = 0.8, col = colorRampPalette(c("blue", "white", "red"))(200))
```

```{r}
# seasonal 
merged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly_filtered %>%
  mutate(Season = case_when(
    Month_UTC %in% c(12, 1, 2) ~ "Winter",
    Month_UTC %in% c(3, 4, 5)  ~ "Spring",
    Month_UTC %in% c(6, 7, 8)  ~ "Summer",
    Month_UTC %in% c(9, 10, 11) ~ "Fall"
  ))

merged_zoop_avg_monthly_filtered$Season <- factor(merged_zoop_avg_monthly_filtered$Season, levels = c("Winter", "Spring", "Summer", "Fall"))

ggplot(merged_zoop_avg_monthly_filtered, aes(x = TA, y = total_plankton)) +
  geom_point(alpha = 0.5, aes(color = Season)) +
  geom_smooth(method = "lm", se = FALSE) +
  scale_color_manual(values = c("Winter" = "#60a5fa",  # Muted Blue
                                "Spring" = "#f9a8d4",  # Muted Pink
                                "Summer" = "#86efac",  # Muted Green
                                "Fall" = "#c08457")) +  
  theme_minimal() +
  facet_wrap(~Season, scales = "free_y") +  # Separate plots for each season
  labs(title = "Effect of DIC on Plankton by Season", 
       x = "DIC(micromoles/kg)", y = "Total Plankton")
```

```{r}
merged_zoop_avg_monthly_filtered <- merged_zoop_avg_monthly_filtered %>% mutate(
  Month_fac = as.factor(Month_UTC)
)
ggplot(merged_zoop_avg_monthly_filtered, aes(x = Season, y = CTDTEMP_ITS90, fill = Season)) +
  geom_boxplot() 
```

```{r}
ggplot(merged_zoop_avg_monthly_filtered, aes(x = Season, y = total_plankton, fill = Season)) +
  geom_boxplot() 
```

